---
title: |
  | Introduction to the Design and Analysis of Randomized Experiments
  | Class 3: Estimators and Power, Randomization-Based Statistical Inference
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: Jake Bowers
bibliography:
 - ../course.bib
fontsize: 10pt
graphics: yes
biblio-style: authoryear-comp
link-citations: true
biblatexoptions:
  - natbib=true
header-includes: |
  \setbeameroption{hide notes}
  \usepackage{hyperref}
  \usepackage{bookmark}
  \setbeamertemplate{footline}{\begin{beamercolorbox}{section in head/foot}
  \insertframenumber/\inserttotalframenumber \end{beamercolorbox}}
  \setbeamertemplate{itemize item}[circle]
  \setbeamertemplate{itemize subitem}{\raisebox{0.2em}{\scalebox{1}{$\blacktriangleright$}}}
  \setbeamersize{description width=2ex}
  \setbeamersize{text margin left=1ex,text margin right=1ex}
  \usepackage{tikz}
  \usepackage{tikz-cd}
  \usepackage{textpos}
  \usepackage{booktabs,multirow,makecell}
  \usepgflibrary{arrows}
  \usetikzlibrary{arrows}
  \tikzstyle{every picture}+=[remember picture]
  \newcommand{\theHtable}{\thetable}
  \usepackage{amsmath,amsthm,bm}
  \DeclareMathOperator{\Var}{\rm{Var}}
  \DeclareMathOperator{\E}{\rm{E}}
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    toc: true
    latex_engine: lualatex
    citation_package: biblatex
    incremental: false
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl", "--toc" ]
colorlinks: true
---

```{r setup, include=FALSE,echo=FALSE}
# Load all the libraries we need
library(here)
library(kableExtra)
library(styler)
library(coin)
library(multcomp)
library(devtools)
library(randomizr)
library(rcompanion) ## for pairwisePermutationTest()
library(DeclareDesign)
library(estimatr)
library(tidyverse)
library(RItools)
```

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## from https://bookdown.org/yihui/rmarkdown-cookbook/font-color.html
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf(
      "<span style='color: %s;'>%s</span>", color,
      x
    )
  } else {
    x
  }
}
```


```{r makedat, echo=FALSE}
## First, create some data,
##  y0 is potential outcome to control
N <- 20
set.seed(123)
dat <- data.frame(
  y0 = c(rpois(N - 2, lambda = 2), rpois(2, lambda = 100))
)
## Just messing around with heterogeneous treatment effects
dat$tau0 <- rnorm(n = N, mean = 0, sd = sd(dat$y0))
dat$tau <- ifelse(dat$tau0 < 0, 0, signif(dat$tau0, 2))
dat$T <- complete_ra(N)
dat$y1 <- with(dat, y0 + tau)
dat$Y <- with(dat, T * y1 + (1 - T) * y0)
dat$Ybin <- as.numeric(dat$Y > median(dat$Y))
## Setup for use with coin
dat$TF <- factor(dat$T)
```



# Overview and Review

##  Today

  0. Quiz and Questions
  1. Estimands, Estimators, Estimation
  2. Statistical Power (1 - false negative rate of tests)

Note: You can download (and contribute to) course materials at [https://github.com/bowers-illinois-edu/short_course_experiments](https://github.com/bowers-illinois-edu/short_course_experiments)

Hay recursos en español aqui: <https://egap.github.io/theory_and_practice_of_field_experiments_spanish/>

## Lingering Questions?

Questions arising?

## Quiz 1

If you worry about chance imbalance on an important covariate, what can you
do **before** randomization to avoid this problem? (For example, in a study of
health outcomes, baseline health. In a study of political activity, baseline
political activity?)

## Quiz 2

We know that we can test the sharp null hypothesis of no effects by
repeating the randomization of the experiment using the `coin` package:

```{r simptest, echo=TRUE}
t_test <- oneway_test(Y ~ TF,
  data = dat,
  distribution = approximate(nresample = 1000), alternative = "less"
)
rank_test <- wilcox_test(Y ~ TF,
  data = dat,
  distribution = approximate(nresample = 1000), alternative = "less"
)
## the 99 percent confidence intervals refer to estimated simulation variability
pvalue(t_test)
pvalue(rank_test)
```

Roughly speaking: how would we do these two tests by hand (they each use the
same randomization design, but different test statistics)?

```{r}
new_experiment <- function(Z){ sample(Z) }
mean_diff_test_stat <- function(Y,Z){ mean(Y[Z==1] - mean(Y[Z==0]))}
mean_diff_rank_test_stat <- function(Y,Z){ranky<-rank(Y); mean(ranky[Z==1] - mean(ranky[Z==0]))}

set.seed(12345)
mean_diff_null_dist <- replicate(1000,mean_diff_test_stat(Y=dat$Y,Z=new_experiment(dat$T)))
set.seed(12345)
mean_diff_rank_null_dist <- replicate(1000,mean_diff_rank_test_stat(Y=dat$Y,Z=new_experiment(dat$T)))

obs_mean_diff_test_stat <- mean_diff_test_stat(Y=dat$Y,Z=dat$T)
obs_mean_diff_rank_test_stat <- mean_diff_rank_test_stat(Y=dat$Y,Z=dat$T)

## two-sided p-values
2*min(mean(mean_diff_null_dist>=obs_mean_diff_test_stat),
  mean(mean_diff_null_dist<obs_mean_diff_test_stat))

2*min(mean(mean_diff_rank_null_dist>=obs_mean_diff_test_stat),
  mean(mean_diff_rank_null_dist<obs_mean_diff_test_stat))
```

## Quiz 3

When would we prefer to test a hypothesis using the direct permutation
 approach, and when would we prefer to use asymptotic approximations?

```{r simptest_asymp, echo=TRUE}
t_test_asymp <- oneway_test(Y ~ TF,
  data = dat,
  distribution = asymptotic(),
)
rank_test_asymp <- wilcox_test(Y ~ TF,
  data = dat,
  distribution = asymptotic(),
)
## the 99 percent confidence intervals refer to estimated simulation variability
pvalue(t_test_asymp)
pvalue(rank_test_asymp)
```

## Quiz 4

Say someone suggested that we test the weak null hypothesis of no effects using
a logit likelihood model (looking at the binary outcome `Ybin`), can you explain
roughly what is going on in the following code?

```{r}
library(broom)

collect_p_values <- function(Y, Z) {
  ## Z is the observed treatment assignment
  ## this next line shuffles Z. Which makes the relationship between Z and Y ...
  ## Make a dataframe for ease but not speed
  newdat <- data.frame(newZ = sample(Z), Y = Y, YF = factor(Y))
  newdat$newZF <- factor(newdat$newZ)

  logit_test <- tidy(glm(Y ~ newZ, data = newdat, family = binomial(link = "logit")))
  p_logit <- logit_test$p.value[logit_test$term == "newZ"]

  ols_test <- tidy(lm(Y ~ newZ, data = newdat))
  p_ols <- ols_test$p.value[ols_test$term == "newZ"]

  ## This next using the randomization justified SE
  neyman_test <- tidy(lm_robust(Y ~ newZ, data = newdat, se_type = "HC2"))
  p_neyman <- neyman_test$p.value[neyman_test$term == "newZ"]
  
  ## Using permutations for the distribution of the difference in means test
  t_test_rand <- oneway_test(Y ~ newZF, data = newdat, distribution = approximate(nresample = 1000))
  p_t_test_rand <- pvalue(t_test_rand)[[1]]

  ## trying a different test since we have a 2x2 table
  ## Strictly speaking this is a test of the shape null and the others are tests of different weak nulls.
  chisq_rand_test <- chisq_test(YF ~ newZF, data = newdat, distribution = approximate(nresample = 1000))
  p_rand_chisq <- pvalue(chisq_rand_test)[[1]]

  chisq_asymp_test <- chisq_test(YF ~ newZF, data = newdat, distribution = asymptotic())
  p_asymp_chisq <- pvalue(chisq_asymp_test)[[1]]

  res <- c(
    p_logit = p_logit,
    p_ols = p_ols,
    p_neyman = p_neyman,
    p_t_test_rand = p_t_test_rand,
    p_asymp_chisq = p_asymp_chisq,
    p_rand_chisq = p_rand_chisq
  )
  return(res)
}
```

```{r}
## Repeat the simulation

many_ps <- replicate(1000, collect_p_values(Y = dat$Ybin, Z = dat$T))

ptils <- c(.05,.1,.2)
res <- apply(many_ps, 1, function(x) {
  ecdf(x)(ptils)
})
cbind(ptils,res)

## Another way to show the same thing
apply(many_ps, 1, function(x) {
  mean(x <= .1)
})
```

# Review of Hypothesis Testing

## Hypothesis testing

 - A hypothesis is a statement about a relationship between potential outcomes 
 - A hypothesis like $Y_i(1)=Y_i(0)$ or $\bar{Y}_i(1) = \bar{Y}_i(0)$ implies a relationship between potential
 outcomes and observed outcomes via treatment assignment, for example: $Y_i = T
 Y_i(1) + (1-T) Y_i(0)$
 - A test statistic summarizes the relationship between $T$ and $Y$ (for example `mean(Y[T==1]) - mean(Y[T==0])`).
 - We can see how this test statistic varies by repeating the design of the experiment.
 - A $p$-value summarizes the relationship between the observed test statistic and the many he hypothetical test statistics that are described by the reference or randomization distribution.
 
 A $p$-value requires a distribution. Where does this distribution come from? What justifies its use?


## Hypothesis testing of the sharp null

 - A sharp null hypothesis is a statement about a relationship between potential outcomes for all units 
 - We can see how this test statistic varies by repeating the design of the experiment.
 - Because we know exactly how the hypothesized outcomes would look, we can **generate** the distribution for the test statistic exactly (or approximate it very closely)

## Hypothesis testing of the weak null {.shrink}

The finite population CLT tells us that

\begin{align*}
\cfrac{\hat{\tau}\left(Z, Y \right) - \E\left[\hat{\tau}\left(Z, Y\right)\right]}{\sqrt{\Var\left[\hat{\tau}\left(Z, Y\right)\right]}} & \overset{d}{\to} \mathcal{N}\left(0, 1\right)
\end{align*}

Diff-in-Means is unbiased, so write

\begin{align*}
\cfrac{\hat{\tau}\left(Z, Y\right) - \tau}{\sqrt{\Var\left[\hat{\tau}\left(Z, Y\right)\right]}} & \overset{d}{\to} \mathcal{N}\left(0, 1\right)
\end{align*}

The CLT is an asymptotic results as $N \to \infty$

Thus, with experiments of at least moderate size and outcomes that aren't too skewed or have extreme outliers,

\begin{align*}
\cfrac{\hat{\tau}\left(Z, Y\right) - \tau}{\sqrt{\Var\left[\hat{\tau}\left(Z, Y\right)\right]}} & \overset{\text{approx.}}{\sim} \mathcal{N}\left(0, 1\right)
\end{align*}

This justifies use of standard Normal distribution for hypothesis tests


# Estimation

## Design Based Approach 2: Estimate Averages of Potential Outcomes

  1. Notice that the observed $Y_i$ are a sample from the (small, finite) population of unobserved potential outcomes $(y_{i,1},y_{i,0})$.
  2. Decide to focus on the average, $\bar{\tau}$, because sample averages, $\hat{\bar{\tau}}$ are unbiased and consistent estimators of population averages.
  3. Estimate $\bar{\tau}$ with the observed difference in means as $\hat{\bar{\tau}}$.

\centering
  \includegraphics[width=.6\textwidth]{../Images/cartoonNeyman.pdf}

## Design Based Approach 2: Estimate Averages of Potential Outcomes

\centering
  \includegraphics[width=.9\textwidth]{../Images/cartoonNeyman.pdf}


## Design Based Approach 2: Estimate Averages of Potential Outcomes

Here using Neyman's standard errors (which @lin_agnostic_2013 shows are the same
as HC2 SEs) and Central Limit Theorem based $p$-values and 95% confidence
intervals:

\smallskip

```{r dim, echo=TRUE}
est1 <- lm_robust(Y ~ T, data = dat)
kable(tidy(est1),digits=3)
## Notice that the estimator is the same as the difference of means
with(dat, mean(Y[T == 1]) - mean(Y[T == 0]))
```

## Key points about estimation I

  -  A causal effect, $\tau_i$, is a comparison of unobserved potential outcomes for each unit $i$: examples  $\tau_{i} = Y_{i}(T_{i}=1) -  Y_{i}(T_{i}=0)$ or  $\tau_{i} = \frac{Y_{i}(T_{i}=1)}{ Y_{i}(T_{i}=0)}$.

  - To learn about $\tau_{i}$, we can treat $\tau_{i}$ as an **estimand** or
   target quantity to be estimated (discussed here) or as a target quantity to be hypothesized about (session on hypothesis testing).

  - Many focus on the **average treatment effect (ATE)**, $\bar{\tau}=\sum_{i=1}^n\tau_{i}$, in part, because it allows for easy **estimation**.

## Key points about estimation II

 - They key to estimation for causal inference is to choose an estimand that
 helps you learn about your theoretical or policy question. So, one could use
   the ATE but other common estimands include the intent to treat effect (ITT),
   the local average treatment effect/complier average causal effect
   (LATE/CACE), average effect on the treated (ATT), or average treatment effect
   (ATE) for some subgroup (or even a difference of causal effects between
   groups).

  - An **estimator** is a recipe for calculating a guess about the value of  an
  estimand. For example, the difference of observed means for $m$ treated units
  is one estimator of $\bar{\tau}$: $\hat{\bar{\tau}} = \frac{\sum_{i=1}^n (T_i
  Y_i)}{m} - \frac{\sum_{i=1}^n ( ( 1 - T_i)Y_i)}{(n-m)}$.

## Key points about estimation III

 - The **standard error** of an estimator in a randomized experiment summarizes how the estimates would vary if the experiment were repeated.

 - We use the **standard error** to produce **confidence intervals** and
   **p-values**: so that we can begin with an estimator and end at a hypothesis test.

  - Different randomizations will produce different values of the same estimator targeting the same estimand. A **standard error** summarizes this variability in an estimator.

  - A $100(1-\alpha)$% **confidence interval** is a collection of hypotheses that cannot be rejected at the $\alpha$ level. We tend to report confidence intervals containing hypotheses about values of our estimand and use our estimator as a test statistic.

## Key points about estimation IV

 - Estimators should:

      - avoid systematic error in their guessing of the estimand (be unbiased);

      - vary little in their guesses from experiment to
   experiment (be precise or efficient); and

      - perhaps ideally converge to the estimand as they use more and more information (be consistent).
      
Notice in our running example the estimated ATE is not the same as the truth:

```{r}

## True ATE
trueATE <- with(dat,mean(y1-y0))
estATE <- coef(est1)[["T"]]
trueATE
estATE

```

## Key points about estimation V

 - **Analyze as you randomize** in the context of estimation means that (1) our
 standard errors should measure variability from randomization and (2) our
 estimators should target estimands defined in terms of potential outcomes.

 - We do not **control for** background covariates when we analyze data from
 randomized experiments. But covariates can make our estimation more
 **precise**. This is called **covariance adjustment** (or covariate
 adjustment). **Covariance adjustment** in randomized experiments differs from
 **controlling for** in observational studies.

## A common estimand and estimator: The average treatment effect and the difference of means

Say we are interested in the ATE, or $\bar{\tau}=\sum_{i=1}^n \tau_{i}$. What is a good estimator?

Two candidates:

 1. The difference of means: $\hat{\bar{\tau}} = \frac{\sum_{i=1}^n (T_i Y_i)}{m} -
    \frac{\sum_{i=1}^n ( ( 1 - T_i) Y_i)}{n-m}$.

 2. A difference of means after top-coding the highest $Y_i$ observation (a
    kind of "winsorized" mean to prevent extreme values from exerting too much
    influence over our estimator --- to increase *precision*).

How would we know which estimator is best for our particular research design?

Let's simulate! (One can **prove** that the first estimator is unbiased. But
simulation is also fun.)

## Simulation Step 1: create some data with a known ATE

Notice that we need to *know* the potential outcomes and the treatment
assignment in order to learn whether our proposed estimator does a good job.
Here we make some new data just for fun.

```{r echo=FALSE}
## First, create some data,
##  y0 is potential outcome to control
N <- 10
y0 <- c(0, 0, 0, 1, 1, 3, 4, 5, 190, 200)
## Each unit has its own treatment effect
tau <- c(10, 30, 200, 90, 10, 20, 30, 40, 90, 20)
## y1 is potential outcome to treatment
y1 <- y0 + tau
## Z is treatment assignment (note we're using Z instead of T)
set.seed(12345)
block <- c("a", "a", "a", "a", "a", "a", "b", "b", "b", "b")
Z <- c(0, 0, 0, 0, 1, 1, 0, 0, 1, 1)
## Y is observed outcomes
Y <- Z * y1 + (1 - Z) * y0
## The data
dat <- data.frame(Z = Z, y0 = y0, y1 = y1, tau = tau, b = block, Y = Y)
```

\begin{center}
```{r}
kbl(dat[, c("Z", "y0", "y1","tau")])
```
\end{center}

```{r ate, echo=FALSE, results="markup", message=TRUE}
ATE <- with(dat, mean(y1 - y0))
message("The true ATE is ", ATE)
```

In reality, we would observe only one of the potential outcomes.

Note that each unit has its own treatment effect.

## First make fake data

The table in the previous slide was generated in R with:


```{r echo=TRUE}
# We have ten units
N <- 10
#  y0 is potential outcome to control
y0 <- c(0, 0, 0, 1, 1, 3, 4, 5, 190, 200)
# Each unit has its own treatment effect
tau <- c(10, 30, 200, 90, 10, 20, 30, 40, 90, 20)
# y1 is potential outcome to treatment
y1 <- y0 + tau
# Two blocks, a and b
block <- c("a", "a", "a", "a", "a", "a", "b", "b", "b", "b")
# Z is treatment assignment (Z instead of T in the code)
Z <- c(0, 0, 0, 0, 1, 1, 0, 0, 1, 1)
# Y is observed outcomes
Y <- Z * y1 + (1 - Z) * y0
# The data
dat <- data.frame(Z = Z, y0 = y0, y1 = y1, tau = tau, b = block, Y = Y)
set.seed(12345)
```


## Using DeclareDesign

Try using DeclareDesign to assess the "goodness" of these two estimators.
DeclareDesign represents research designs in a few steps shown below:

```{r dd1, echo=TRUE}
# take just the potential outcomes under treatment and control from our
# fake data
small_dat <- dat[, c("y0", "y1")]

# DeclareDesign first asks you to declare your population
pop <- declare_population(small_dat)

# 5 units assigned to treatment; default is simple random assignment with
# probability 0.5
trt_assign <- declare_assignment(Z = complete_ra(N = nrow(small_dat), m = 5))

# observed Y is y1 if Z=1 and y0 if Z=0
pot_out <- declare_potential_outcomes(Y ~ Z * y1 + (1 - Z) * y0)

# specify outcome and assignment variables
reveal <- declare_reveal(Y, Z)

# the basic research design object includes these four objects
base_design <- pop + trt_assign + pot_out + reveal
```

## Using DeclareDesign: make fake data

DeclareDesign renames `y0` and `y1` by default to `Y_Z_0` and `Y_Z_1`:

```{r echo=TRUE}
## A simulation is one random assignment of treatment
sim_dat1 <- draw_data(base_design)

## Simulated data (just the first 6 lines)
head(sim_dat1)
```

## Using DeclareDesign: define estimand and estimators

No output here. Just define functions and estimators and one estimand.

```{r dd2, echo=TRUE}
## The estimand
estimandATE <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))

## The first estimator is difference-in-means
diff_means <- declare_estimator(Y ~ Z,
  inquiry = estimandATE,
  .method = lm_robust, se_type = "classical", label = "Diff-Means/OLS"
)
```

## Using DeclareDesign: define estimand and estimators
```{r dd2a, echo=TRUE}
## The second estimator is top-coded difference-in-means
diff_means_topcoded_fn <- function(data) {
  data$rankY <- rank(data$Y)
  ## Code the maximum value of Y as the second to maximum value of Y
  data$newY <- with(
    data,
    ifelse(rankY == max(rankY), Y[rankY == (max(rankY) - 1)], Y)
  )
  obj <- lm_robust(newY ~ Z, data = data, se_type = "classical")
  res <- tidy(obj) %>% filter(term == "Z")
  return(res)
}
diff_means_topcoded <- declare_estimator(
  handler = label_estimator(diff_means_topcoded_fn),
  inquiry = estimandATE, label = "Top-coded Diff Means"
)
```

## Using DeclareDesign: define estimand and estimators

Here we show how the DD estimators work using our simulated data.

```{r dd3, echo=TRUE}
## Demonstrate that the estimand works:
estimandATE(sim_dat1)

## Demonstrate that the estimators estimate
## Estimator 1 (difference in means)
diff_means(sim_dat1)[-c(1, 2, 10, 11)]

## Estimator 2 (top-coded difference in means)
diff_means_topcoded(sim_dat1)[-c(1, 2, 10, 11)]
```


## Then simulate with one randomization

Recall the true ATE:

```{r trueATE, echo=TRUE}
trueATE <- with(sim_dat1, mean(y1 - y0))
with(sim_dat1, mean(Y_Z_1 - Y_Z_0))
```

In one experiment (one simulation of the data) here are the simple estimates:

```{r echo=TRUE}
## Two ways to calculate the difference of means estimator
est_diff_means_1 <- with(sim_dat1, mean(Y[Z == 1]) - mean(Y[Z == 0]))
est_diff_means_2 <- coef(lm_robust(Y ~ Z,
  data = sim_dat1,
  se = "classical"
))[["Z"]]
c(est_diff_means_1, est_diff_means_2)
```

## Then simulate with one randomization

In one experiment (one simulation of the data) here are the estimates after top-coding:

```{r echo=TRUE}
## Two ways to calculate the topcoded difference of means estimator
sim_dat1$rankY <- rank(sim_dat1$Y)
sim_dat1$Y_tc <- with(sim_dat1, ifelse(rankY == max(rankY),
  Y[rankY == (max(rankY) - 1)], Y
))
est_topcoded_1 <- with(sim_dat1, mean(Y_tc[Z == 1]) - mean(Y_tc[Z == 0]))
est_topcoded_2 <- coef(lm_robust(Y_tc ~ Z,
  data = sim_dat1,
  se = "classical"
))[["Z"]]
c(est_topcoded_1, est_topcoded_2)
```


## Then simulate a different randomization and estimate the ATE with the same estimators

Now calculate your estimate with the same estimators using a  **different** randomization. Notice that the answers differ. The estimators are estimating the *same estimand* but now they have a different randomization to work with.

```{r echo=TRUE}
# do another random assignment of the treatment in DeclareDesign
# this produces a new simulated dataset with a different random assignment
sim_dat2 <- draw_data(base_design)
# the first estimator (difference in means)
coef(lm_robust(Y ~ Z, data = sim_dat2, se = "classical"))[["Z"]]
# the second estimator (top-coded difference in means)
sim_dat2$rankY <- rank(sim_dat2$Y)
sim_dat2$Y_tc <- with(sim_dat2, ifelse(rankY == max(rankY),
  Y[rankY == (max(rankY) - 1)], Y
))
coef(lm_robust(Y_tc ~ Z, data = sim_dat2, se = "classical"))[["Z"]]
```


## How do our estimators behave in general for this design?

Our estimates vary across randomizations. Do our two estimators vary in the same ways?

```{r diagnose, echo=TRUE, cache=TRUE}
## Combine into one DeclareDesign design object
## This has the base design, estimand, then our two estimators
design_plus_ests <- base_design + estimandATE + diff_means +
  diff_means_topcoded
## Run 100 simulations (reassignments of treatment) and
## apply the two estimators (diff_means and diff_means_topcoded)
diagnosis1 <- diagnose_design(design_plus_ests,
  bootstrap_sims = 0, sims = 1000
)
sims1 <- get_simulations(diagnosis1)
head(sims1[, -c(1:6)])
```

## How do our estimators behave in general for this design?

Our estimates vary across randomizations. Do our two estimators vary in the same ways?
  How should we interpret this plot?

```{r sim_plot, out.width=".8\\textwidth"}
sim_plot <- ggplot(sims1, aes(y = estimate, x = estimator, color = estimator)) +
  geom_boxplot() +
  geom_hline(yintercept = trueATE) +
  geom_point(aes(group = estimator)) +
  theme(text = element_text(size = 20))
print(sim_plot)
```


## Which estimator is closer to the truth?

One way to choose among estimators is to choose the one that is **close to the truth** whenever we use it --- regardless of the specific randomization.

An "unbiased" estimator is one for which **average of the estimates across repeated designs** is the  same as the  truth (or $E_R(\hat{\bar{\tau}})=\bar{\tau}$). An unbiased estimator has "no systematic error" but doesn't guarantee closeness to the truth.

Another measure of closeness is **root mean squared error** (RMSE) which  records  squared distances  between the truth and the individual estimates.

Which estimator is better? (One is closer to the truth on average (RMSE) and is more precise. The other has no systematic error --- is unbiased.)

```{r}
kableExtra::kable(reshape_diagnosis(diagnosis1, select = c("Estimator", "Bias", "RMSE", "SD Estimate", "Power"))[, -c(1, 2, 4, 5, 6)])
```


## Unbiased and biased estimators

Summary:

 - We have a *choice* of both estimands and estimators

 - A good estimator performs well regardless of the particular randomization of a given design. And *performs well* can mean "unbiased" and/or "low mse" (or "consistent" --- which means increasingly close to the truth as the sample size increases).

 - We can learn about how a given estimator performs in a given study using simulation.

# Estimation given Block randomization

## Block-randomized experiments are a collection of mini-experiments

What is the **ATE** estimand in a block-randomized experiment?

If we think of the unit-level ATE as: $(1/N) \sum_{i=1}^N y_{i,1} - y_{i,0}$
then we could re-express this equivalently using the ATE in block $j$ is $ATE_j$
as follows:

\begin{equation}
ATE = \frac{1}{J}\sum^J_{j=1} \sum^{N_j}_{i=1} \frac{y_{i,1} - y_{i,0}}{N_j}  = \sum^J_{j=1} \frac{N_j}{N} ATE_j
\end{equation}

And it would be natural to *estimate* this quantity by plugging in what we can calculate:

\begin{equation}
\widehat{ATE} = \sum^J_{j=1} \frac{N_j}{N} \widehat{ATE}_j
\end{equation}


## Block-randomized experiments are a collection of mini-experiments

And we could *define* the standard error of the estimator by also just averaging the within-block standard errors (if our blocks are large enough):

$SE(\widehat{ATE}) = \sqrt{\sum^J_{j=1} (\frac{N_{j}}{N})^2SE^2(\widehat{ATE}_j)}$


## Estimating the ATE in block-randomized experiments

One approach to estimation simply replaces $ATE_j$ with $\widehat{ATE}$ above:

```{r br1, echo=TRUE}
with(dat, table(b, Z))
```

We have 6 units in block `a`, 2 of which are assigned to treatment, and 4 units in block `b`, 2 of which are assignment to treatment.

## Estimating the ATE in block-randomized experiments

One approach to estimation simply replaces $ATE_j$ with $\widehat{ATE}$ above:

```{r br2, echo=TRUE}
datb <- dat %>%
  group_by(b) %>%
  summarize(
    nb = n(), pb = mean(Z), estateb = mean(Y[Z == 1]) - mean(Y[Z == 0]),
    ateb = mean(y1 - y0), .groups = "drop"
  )
datb
## True ate by block:
with(dat, mean(y1 - y0))
## This is another way to calculate the true ate
with(datb, sum(ateb * (nb / sum(nb))))
```


## Estimating the ATE in block-randomized experiments

One approach is to estimate the overall ATE using block-size weights:

```{r br3, echo=TRUE}
## Showing that difference_in_means uses the blocksize weight.
e1 <- difference_in_means(Y ~ Z, blocks = b, data = dat)
e2 <- with(datb, sum(estateb * (nb / sum(nb))))
c(coef(e1)[["Z"]], e2)
```


## Estimating the ATE in block-randomized experiments

Notice that this is **not** the same as either of the following:

```{r br4, echo=TRUE}
## Ignoring blocks
e3 <- lm(Y ~ Z, data = dat)
coef(e3)[["Z"]]

## With block fixed effects
e4 <- lm(Y ~ Z + block, data = dat)
coef(e4)[["Z"]]
```

How do they differ? (The first ignores the blocks. The second uses a different set of weights that are created by use of "fixed effects" or "indicator" or "dummy" variables.)

## Which estimator should we use?

We now have three estimators each with a different estimate (imagining they all target the same estimand):

```{r echo=TRUE}
c(coef(e1)[["Z"]], coef(e3)[["Z"]], coef(e4)[["Z"]])
```

Which estimator should we use for this design?  We can set up a DeclareDesign simulation to figure this out.


```{r blockdd0, cache=TRUE, echo=TRUE}
## declare a new base design that includes the block indicator b
base_design_blocks <-
  # declare the population
  declare_population(dat[, c("b", "y0", "y1")]) +
  # tell DD that b indicates block and to assign 2 treated units in each block
  declare_assignment(
    Z = block_ra(m = 2, blocks = b),
    Z_cond_prob = obtain_condition_probabilities(Z, declaration = declare_ra(m = 2, blocks = b))
  ) +
  # relationship of potential outcomes to observed outcome
  declare_potential_outcomes(Y ~ Z * y1 + (1 - Z) * y0) +
  # observed outcome and treatment assignment
  declare_reveal(Y, Z)
```

## Which estimator should we use?


```{r blockdd1, echo=TRUE, cache=TRUE}
# the estimand is the average treatment effect
estimandATEb <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))

# three different estimators
est1 <- declare_estimator(Y ~ Z,
  inquiry = estimandATEb, .method = lm_robust,
  label = "Ignores Blocks"
)
est2 <- declare_estimator(Y ~ Z,
  inquiry = estimandATEb, .method = difference_in_means, blocks = b,
  label = "DiM: Block-Size Weights"
)
est3 <- declare_estimator(Y ~ Z,
  inquiry = estimandATEb, .method = lm_robust,
  weights = (Z / Z_cond_prob) + ((1 - Z) / (Z_cond_prob)),
  label = "LM: Block Size Weights"
)
```

## Which estimator should we use?


```{r blockdd1a, echo=TRUE, cache=TRUE}
# two more estimators
est4 <- declare_estimator(Y ~ Z,
  inquiry = estimandATEb,
  .method = lm_robust, fixed_effects = ~b, label = "Precision Weights"
)
est5 <- declare_estimator(Y ~ Z + b,
  inquiry = estimandATEb,
  .method = lm_robust, label = "Precision Weights (LSDV)"
)

## new design object has the base design, the estimand, and five estimators
design_blocks <- base_design_blocks + estimandATEb +
  est1 + est2 + est3 + est4 + est5
```

Then we will run 10,000 simulations (reassign treatment 10,000 times) and summarize the estimates produced by each of these five estimators.

## Which estimator should we use?

```{r futurelibs, echo=FALSE}
## This to enable parallel processing when doing the diagnoses.
## Notice that we have turned on "cache" for some code chunks so that they are not
## re-run each time we change the slides.
library(future)
library(future.apply)
```

```{r diagnosis2, echo=FALSE, cache=TRUE}
## The next few lines use all of the cores on your computer to speed up the computation

# plan(multicore)
set.seed(12345)
diagnosis2 <- diagnose_design(design_blocks, bootstrap_sims = 0, sims = 1000)
sims2 <- get_simulations(diagnosis2)
# plan(sequential)
```

How should we interpret this plot?

```{r sim_plot2, warning=FALSE, message=FALSE, out.width=".9\\textwidth"}
sim_plot2 <- ggplot(sims2, aes(y = estimate, x = estimator, color = estimator)) +
  geom_boxplot() +
  geom_hline(yintercept = trueATE) +
  geom_point(aes(group = estimator)) +
  theme(
    text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none", legend.title = element_blank()
  ) +
  xlab("")
print(sim_plot2)
```


## Which estimator is closer to the truth?

Which estimator works better on this design and these data?


```{r blocktab}
blocktab <- reshape_diagnosis(diagnosis2) %>% select(one_of(c("Estimator", "Bias", "RMSE", "SD Estimate", "Power", "Coverage")))
kableExtra::kable(blocktab)
```

Notice that the coverage is not always at 95% in all cases.  We used 10,000
simulations so simulation error is around $\pm 2 \sqrt{p(1-p)/10000}$ or, say,
for coverage calculated as .93, a different simulation could have easily
produced `r round(.93-2*sqrt( .93 * (1-.93)/10000),3 )` or  `r round(.93+2*sqrt( .93 *
(1-.93)/10000),3 )` (or would rarely have produced coverage numbers outside that
range just by chance).

# Estimation given Cluster randomization

## In cluster-randomized experiments, units are randomized as a group (cluster) to treatment  {.allowframebreaks}

- **Example 1:** an intervention is randomized across neighborhoods, so **all** households in a neighborhood will be assigned to the same treatment condition, but different neighborhoods will be assigned different treatment conditions.
- **Example 2:** an intervention is randomized across people and each person is measured four times after treatment, so our data contain four rows per person.
- **Not An Example 1:** Neighborhoods are chosen for the study. Within each neighborhood about half of the people are assigned to treatment and half to control. (What kind of study is this? It is not a cluster-randomized study.)
- **Not an Example 2:** an intervention is randomized to some neighborhoods and not to others, the outcomes include measurements of neighborhood-level trust in government and total land area in the neighborhood devoted to gardens. (Sometimes a cluster randomized experiment can be turned into a simple randomized experiment. Or may contain more than one possible approach to analysis and interpretation.)

How might the distribution of test statistics and estimators differ from an experiment where individual units (not clusters) are randomized?

## Estimating the ATE in cluster-randomized experiments

Bias problems in cluster-randomized experiments:

- When clusters are the same size, the usual difference-in-means estimator is unbiased.

- But be careful when clusters have different numbers of units or you have very few clusters because then treatment effects may be correlated with cluster size.

- When cluster size is related to potential outcomes, the usual difference-in-means estimator is biased. <https://declaredesign.org/blog/bias-cluster-randomized-trials.html>

## Estimating the SE for the ATE in cluster-randomized experiments  {.allowframebreaks}

- **Misleading statistical inferences:** The default SE will generally underestimate precision in such designs and thus produce tests with false positive rates that are too high (or equivalently confidence intervals coverage rates that are too low).

- The "cluster robust standard errors" implemented in common software work well **when the number of clusters is large** (like more than 50 in some simulation studies).

- The default cluster-appropriate standard errors in `lm_robust` (the `CR2` SEs) work better than the common approach in Stata (as of this writing).
- The wild bootstrap helps control error rates but gives up statistical power much more than perhaps necessary in a cluster randomized study where direct randomization inference is possible.

- When in doubt, one can produce $p$-values by direct simulation (direct randomization inference) to see if they agree with one of the cluster robust approaches.

Overall, it is worth simulating to study the performance of your estimators, tests, and confidence intervals if you have any worries or doubts.

## An example of estimation

```{r makedatclus, echo=FALSE, results="hide"}
## see https://declaredesign.org/blog/bias-cluster-randomized-trials.html
## for more like this.
N_clusters <- 10 # number of clusters
n_indivs <- c(100, 10) # possible size of clusters
thepop <- declare_population(
  clus_id = add_level(
    # set the number of clusters
    N = N_clusters,
    # 1/5 of clusters have 100 individuals, 4/5 of clusters have 10 individuals
    cl_size = rep(n_indivs, c(N / 5, N - N / 5)),
    cl_sizeF = factor(cl_size),
    # Each cluster has a different mean level (u) and different background variability (sd of u)
    effect = ifelse(cl_size == 100, .1, 1)
  ),
  indiv = add_level(
    N = cl_size,
    u = rnorm(N, mean = log(cl_size), sd = effect)
  )
)

theys <- declare_potential_outcomes(Y_Z_0 = u, Y_Z_1 = Y_Z_0 + effect)

thetarget_indiv <- declare_inquiry(ATE_indiv = mean(Y_Z_1 - Y_Z_0))

## Complete random assignment to clusters
theassign <- declare_assignment(Z = conduct_ra(N = N, clusters = clus_id))

thereveal <- declare_reveal(Y, Z)

## 7 different estimators
est1 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv", clusters = clus_id,
  .method = lm_robust,se_type="CR2", label = "Y~Z, CR2 SE"
)

est2 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv",
  .method = lm_robust,se_type="HC2", label = "Y~Z, HC2 SE"
)

est3 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv",
  .method = lm_robust, se_type = "classical", label = "Y~Z, IID SE"
)

est4 <- declare_estimator(Y ~ Z + cl_sizeF,
  inquiry = "ATE_indiv", clusters = clus_id,
  .method = lm_robust, se_type="CR2", label = "Y~Z+clus_size_fixed_effects, CR2 SE"
)

est5 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv", fixed_effects = ~cl_sizeF, clusters = clus_id,
  .method = lm_robust, se_type="CR2", label = "Y~Z, fixed_effects=~clus_size_fixed_effects, CR2 SE "
)

est6 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv", covariates = ~cl_size, clusters = clus_id,
  .method = lm_lin, se_type="CR2", label = "Y~Z*I(clus_size-mean(clus_size)), CR2 SE"
)

est7 <- declare_estimator(Y ~ Z,
  inquiry = "ATE_indiv", weight = cl_size, clusters = clus_id,
  .method = lm_robust, se_type="CR2", label = "Y~Z, weight=clus_size, CR2 SE"
)

### Some experimental stuff here:
## remotes::install_github("markmfredrickson/RItools",ref="proj1-balT")
## est7tmp <-  balanceTest(Z~Y+cluster(clus_id),data=dat1,report="all")
##
## est7fn <- function(data){
##     bt <- balanceTest(Z~Y+cluster(clus_id),data=data,report="all")
##     resdat <- data.frame(estimate=bt$results[,"adj.mean diff",])
##     return(resdat)
## }
## est7fn <- function(data){
##     thelme <- lmer(Y~Z+(1|clus_id),data=data)
##     cilme <- confint(thelme)
##     lmecoef <- summary(thelme)$coefficients["Z",]
##     resdat <- data.frame(estimate=lmecoef["Estimate"],
##         std.error=lmecoef["Std. Error"],
##             statistic=lmecoef["t value"],
##             p.value=NA,
##         conf.low=min(cilme["Z",]),
##         conf.high=max(cilme["Z",]))
##     return(resdat)
## }
##
## est7 <- declare_estimator(handler=tidy_estimator(est7fn),label="mlm: rand intercept")

des <- thepop + theys + theassign + thereveal

set.seed(12345)
dat1 <- draw_data(des)

head(dat1)

table(dat1$clus_id)
with(dat1, table(clus_id, Z))
dat1 %>%
  group_by(clus_id) %>%
  summarize(mean(Y_Z_1 - Y_Z_0))

## g1 <- ggplot(data=dat1,aes(x=Y,group=clus_id,fill=clus_id,color=clus_id))+
##     geom_density()
## g1

est1(dat1)
est2(dat1)
est3(dat1)
est4(dat1)
est5(dat1)
est6(dat1)
est7(dat1)
```

Imagine we had data from 10 clusters with either 100 people (for 2 clusters) or 10 people per cluster (for 8 clusters). The total size of the data is `r nrow(dat1)`.


```{r}
tmp <- dplyr::filter(dat1, clus_id %in% c("03", "01")) %>%
  group_by(clus_id) %>%
  sample_n(3) %>%
  arrange(clus_id, indiv) %>%
  dplyr::select(clus_id, indiv, Y_Z_0, Y_Z_1, Z, Y)

tmp
```

## An example of estimation

Which estimator should we use? Which test should we use? On what basis should we
choose among these approaches? (Notice here same estimate but very different
standard errors and thus $p$-values.)

```{r clusest, echo=TRUE}
options(digits=4)
lmc1 <- lm_robust(Y ~ Z, data = dat1)
lmc2 <- lm_robust(Y ~ Z, clusters = clus_id, data = dat1)
lmc3 <- lm_robust(Y ~ Z + cl_sizeF, clusters = clus_id, data = dat1)
tidy(lmc1)[2, ]
tidy(lmc2)[2, ]
tidy(lmc3)[2, ]
```


## Use simulation to assess estimators and tests

If you look at the code for the slides you will see that we simulate the design
5000 times, each time calculating an estimate and confidence interval for
different estimators of the ATE.

What should we learn from this table? (Coverage? `sd_estimate` versus `mean_se`).

```{r simdesign, warning=FALSE, results="hide",message=FALSE}
des_plus_est <- des + thetarget_indiv + est1 + est2 + est3 + est4 + est5 +  est7
des_plus_est
draw_estimates(des_plus_est)
```

```{r diag_clust, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(12346)
plan(multicore)
diag_clus <- diagnose_design(des_plus_est, bootstrap_sims = 0, sims = 1000)
sim_clus <- get_simulations(diag_clus) # simulate_design(des_plus_est,sims=1000)
trueclusATE <- thetarget_indiv(dat1)[["estimand"]]
plan(sequential)
```


```{r cluster_sim_res}
res_clus <- sim_clus %>%
  na.omit() %>%
  group_by(estimator) %>%
  summarize(
    bias = mean(estimate - estimand),
    rmse = sqrt(mean((estimate - estimand)^2)),
    power = mean(p.value < .05),
    coverage = mean(estimand <= conf.high & estimand >= conf.low),
    # mean_estimate = mean(estimate),
    sd_estimate = sd(estimate),
    mean_se = mean(std.error)
  )


#res_clus[2, "estimator"] <- "Y~Z, cl_size fe, CR2"
#res_clus[6, "estimator"] <- "Y~Z*I(cl_size-mean(cl_size)), CR2"
#res_clus[7, "estimator"] <- "Y~Z+cl_sizeF, CR2"
res_clus$estimator <- gsub(" SE", "", res_clus$estimator)
```

```{r showresclus1}
kableExtra::kable(res_clus[, c(1, 5:7)], digits = 2, booktabs = TRUE, linesep = "", caption = "Estimator and Test Performance in 5000 simulations of the cluster randomized design for different estimators and confidence intervals")
```

## Use simulation to assess estimators and tests

What should we learn from this table? (Bias? Closeness to truth?)

```{r showresclus2}
kableExtra::kable(res_clus[, c(1:3)], digits = 3, booktabs = TRUE, linesep = "", caption = "Estimator and Test Performance in 5000 simulations of the cluster randomized design for different estimators and confidence intervals")
```

## Use simulation to assess estimators and tests

How should we interpret this plot?

```{r sim_plot_clus, warning=FALSE, out.width=".95\\textwidth"}
sim_plot3 <- ggplot(sim_clus, aes(y = estimate, x = estimator, color = estimator)) +
  geom_boxplot() +
  coord_flip() +
  geom_hline(yintercept = trueclusATE) +
  geom_point(aes(group = estimator)) +
  theme(
    text = element_text(size = 20),
    legend.position = "none", legend.title = element_blank()
  ) +
  ylab("")
print(sim_plot3)
```

## Summary of estimation and testing in cluster-randomized trials

 - Cluster randomized trials pose special problems for standard approaches to estimation and testing.

 - If randomization is at the cluster level, then uncertainty arises from the cluster level randomization.

 - If we have enough clusters, then one of the "cluster robust" standard errors can help us produce confidence intervals with correct coverage. **Cluster robust standard errors require many clusters**.

 - If cluster size (or characteristic) is related to effect size, then we can have bias (and we need to adjust somehow).


# Binary outcomes

## Binary outcomes: Set up our data for simulation in DeclareDesign
```{r setupbin, echo=TRUE}
# population size
N <- 20
# declare the population
thepop_bin <- declare_population(
  N = N, x1 = draw_binary(prob = .5, N = N),
  x2 = rnorm(N)
)
# declare the potential outcomes
thepo_bin <- declare_potential_outcomes(Y ~ rbinom(
  n = N, size = 1,
  prob = 0.5 + 0.05 * Z + x1 * .05
))
# two possible targets: difference in means or difference in log-odds
thetarget_ate <- declare_inquiry(ate = mean(Y_Z_1 - Y_Z_0))
thetarget_logodds <- declare_inquiry(
  logodds = log(mean(Y_Z_1) / (1 - mean(Y_Z_1))) -
    log(mean(Y_Z_0) / (1 - mean(Y_Z_0)))
)
```

## Binary outcomes: Set up our data for simulation in DeclareDesign
```{r setupbin2, echo=TRUE}
# declare how treatment is assigned
# m units are assigned to levels of treatment Z
theassign_bin <- declare_assignment(Z = conduct_ra(N = N, m = floor(N / 3)))
# declare what outcome values are revealed for possible values of Z
thereveal_bin <- declare_reveal(Y, Z)
# pull this all together: population, potential outcomes, assignment,
## outcome values connected to Z
des_bin <- thepop_bin + thepo_bin + theassign_bin + thereveal_bin
# then make one draw (randomize treatment once)
set.seed(12345)
dat2 <- draw_data(des_bin)
```

## Binary outcomes: Estimands I

How would we interpret the following true
quantities or estimands? (`Y_Z_1`, `Y_Z_0` are potential
outcomes, `Y` is observed, `x1`, `x2` are covariates, `Z` is treatment assignment. Here $N$=`r nrow(dat2)`.

```{r dat2echo, echo=TRUE}
## Look at the first 6 observations only:
head(dat2[, -7])
```


## Binary outcomes: Estimands II

How would we interpret the following true
quantities or estimands? (`Y_Z_1`, `Y_Z_0` are potential
outcomes, `Y` is observed, `x1`, `x2` are covariates, `Z` is treatment assignment. Here $N$=`r nrow(dat2)`.

```{r bin1, echo=TRUE}
ate_bin <- with(dat2, mean(Y_Z_1 - Y_Z_0))
bary1 <- mean(dat2$Y_Z_1)
bary0 <- mean(dat2$Y_Z_0)
diff_log_odds_bin <- with(
  dat2,
  log(bary1 / (1 - bary1)) - log(bary0 / (1 - bary0))
)
c(
  bary1 = bary1, bary0 = bary0, true_ate = ate_bin,
  true_diff_log_odds = diff_log_odds_bin
)
```

## Binary outcomes: Estimands III

Do you want to estimate the difference in log-odds?

\begin{equation}
\delta = \log \frac{\bar{y}_{1}}{1-\bar{y}_{1}} - \log \frac{ \bar{y}_0}{1- \bar{y}_0}
\end{equation}

Or the difference in proportions?

\begin{equation}
\bar{\tau} = \bar{y}_{1} - \bar{y}_0
\end{equation}

Recall that $\bar{y}_1$ is the *proportion* of $y_{1}=1$ in the data.

@freedman2008randomization shows us that the logit coefficient estimator is a biased estimator of the difference in log-odds estimand. He also shows an unbiased estimator of that estimand.

We know that the difference of proportions in the sample should be an unbiased estimator of the difference of proportions.


## An example of estimation I

How should we interpret the following estimates? (What does the difference of
means estimator require in terms of assumptions? What does the logistic
regression estimator require in terms of assumptions?)

```{r estexample, echo=TRUE}
lmbin1 <- lm_robust(Y ~ Z, data = dat2)
glmbin1 <- glm(Y ~ Z, data = dat2, family = binomial(link = "logit"))
library(broom)
tidy(lmbin1)[2, ]
tidy(glmbin1)[2, ]
```

## An example of estimation II

What about with covariates?  Why use covariates?

```{r estexample2, echo=TRUE}
lmbin2 <- lm_robust(Y ~ Z + x1, data = dat2)
glmbin2 <- glm(Y ~ Z + x1, data = dat2, family = binomial(link = "logit"))

tidy(lmbin2)[2, ]
tidy(glmbin2)[2, ]
```

## An example of estimation III

Let's compare our estimates

```{r estexample3, echo=TRUE}
c(
  dim = coef(lmbin1)[["Z"]],
  dim_x1 = coef(lmbin2)[["Z"]],
  glm = coef(glmbin1)[["Z"]],
  glm_x1 = coef(glmbin2)[["Z"]]
)
```

## An example of estimation: The Freedman plugin estimators I

No covariate:
```{r pluginest, echo=TRUE }
freedman_plugin_estfn1 <- function(data) {
  glmbin <- glm(Y ~ Z, data = dat2, family = binomial(link = "logit"))
  preddat <- data.frame(Z = rep(c(0, 1), nrow(dat2)))
  preddat$yhat <- predict(glmbin, newdata = preddat, type = "response")
  bary1 <- mean(preddat$yhat[preddat$Z == 1])
  bary0 <- mean(preddat$yhat[preddat$Z == 0])
  diff_log_odds <- log(bary1 / (1 - bary1)) - log(bary0 / (1 - bary0))
  return(data.frame(estimate = diff_log_odds))
}
```

## An example of estimation: The Freedman plugin estimators II

With covariate:
```{r pluginest2, echo=TRUE }
freedman_plugin_estfn2 <- function(data) {
  N <- nrow(data)
  glmbin <- glm(Y ~ Z + x1, data = data, family = binomial(link = "logit"))
  preddat <- data.frame(Z = rep(c(0, 1), each = N))
  preddat$x1 <- rep(data$x1, 2)
  preddat$yhat <- predict(glmbin, newdata = preddat, type = "response")
  bary1 <- mean(preddat$yhat[preddat$Z == 1])
  bary0 <- mean(preddat$yhat[preddat$Z == 0])
  diff_log_odds <- log(bary1 / (1 - bary1)) - log(bary0 / (1 - bary0))
  return(data.frame(estimate = diff_log_odds))
}
```

Let's compare our estimates from the six different estimators
```{r echo=FALSE}
c(
  dim = coef(lmbin1)[["Z"]],
  dim_x1 = coef(lmbin2)[["Z"]],
  glm = coef(glmbin1)[["Z"]],
  glm_x1 = coef(glmbin2)[["Z"]],
  freedman = freedman_plugin_estfn1(dat2)[["estimate"]],
  freeman_x1 = freedman_plugin_estfn2(dat2)[["estimate"]]
)
```


```{r tmleapproach, eval=FALSE}
## Here is another approach to using the plugin estimator but allowing for standard errors, etc..
## Doesn't require a handwritten function like the ones used above.
library(tmle)
Y <- as.matrix(dat2$Y, ncol = 1)
A <- as.matrix(dat2$Z, ncol = 1)
W <- as.matrix(dat2[, c("x1", "x2")], ncol = 1)
colnames(W) <- paste("W", 1:2, sep = "")
tmle1 <- tmle(
  Y = Y, A = A, W = W,
  family = "binomial", Qform = Y ~ A, gform = A ~ 1, cvQinit = FALSE,
  Q.SL.library = "SL.glm", g.SL.library = "SL.glm", g.Delta.SL.library = "SL.glm"
)

tmle1$estimates$ATE
tmle1$estimates$OR

tmle2 <- tmle(
  Y = Y, A = A, W = W,
  family = "binomial", Qform = Y ~ A + W1, gform = A ~ 1, cvQinit = FALSE, # V=0,
  Q.SL.library = "SL.glm", g.SL.library = "SL.glm", g.Delta.SL.library = "SL.glm",
  obsWeights = NULL
)

tmle2$estimates$ATE
tmle2$estimates$OR
```

## An example of using DeclareDesign to assess our estimators I

```{r ddbinsetup, echo=TRUE}
# declare 4 estimators for DD
# first estimator: linear regression with ATE as target
estb1 <- declare_estimator(Y ~ Z,
  .method = lm_robust, label = "lm1:Z",
  inquiry = thetarget_ate
)
# second estimator: linear regression with covariate, with ATE as target
estb2 <- declare_estimator(Y ~ Z + x1,
  .method = lm_robust, label = "lm1:Z,x1",
  inquiry = thetarget_ate
)
# third estimator: logistic regression, with log odds as target
estb3 <- declare_estimator(Y ~ Z,
  .method = glm, family = binomial(link = "logit"),
  label = "glm1:Z", inquiry = thetarget_logodds
)
# fourth estimtor: logistic regression with covariate, with log odds as target
estb4 <- declare_estimator(Y ~ Z + x1,
  .method = glm, family = binomial(link = "logit"),
  label = "glm1:Z,x1", inquiry = thetarget_logodds
)
```


## An example of using DeclareDesign to assess our estimators II

```{r ddbinsetup2, echo=TRUE}
# Pull together: des_bin is population, potential outcomes, assignment,
# outcome values connected to Z.  We add the two targets and four estimators.
des_bin_plus_est <- des_bin + thetarget_ate + thetarget_logodds +
  estb1 + estb2 + estb3 + estb4
```

```{r diagnosis_bin, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
## The next few lines use all of the cores on your computer to speed up the computation
library(future)
library(future.apply)
# plan(multiprocess)
set.seed(12345)
diagnosis_bin <- diagnose_design(des_bin_plus_est, bootstrap_sims = 0, sims = 1000)
sims_bin <- get_simulations(diagnosis_bin)
trueATE_bin <- thetarget_ate(dat2)[["estimand"]]
truelo_bin <- thetarget_logodds(dat2)[["estimand"]]
# plan(sequential)
```

## Using simulation to assess our estimators

How should we interpret this plot? (Differences in scales make it difficult.)

```{r sim_plot_bin, out.width=".95\\textwidth"}
estimand_dat <- sims_bin %>%
  group_by(inquiry) %>%
  summarize(meanestimand = mean(estimand))
sim_plot_bin <- ggplot(sims_bin, aes(y = estimate, x = estimator, color = estimator)) +
  geom_boxplot() +
  geom_point() +
  facet_wrap(~inquiry, scales = "free") +
  geom_hline(data = estimand_dat, aes(yintercept = meanestimand)) +
  theme(
    text = element_text(size = 20),
    legend.position = "none", legend.title = element_blank()
  )
print(sim_plot_bin)
```

## Which estimator is closer to the truth?

Which estimator works better on this design and these data?

```{r bin_sim_res, message=FALSE,warning=FALSE}
## Notice that the lin estimator is great but sometimes cannot produce an answer
res_bin <- sims_bin %>%
  group_by(estimator, inquiry) %>%
  summarize(
    bias = mean(estimate - estimand),
    rmse = sqrt(mean((estimate - estimand)^2)),
    power = mean(p.value < .05),
    coverage = mean(estimand <= conf.high & estimand >= conf.low, na.rm = TRUE),
    # mean_estimate = mean(estimate),
    sd_est = sd(estimate),
    mean_se = mean(std.error)
  )
names(res_bin)[1:2] <- c("est", "estimand")
```

```{r showresbin1}
kableExtra::kable(res_bin, digits = 3, booktabs = TRUE, linesep = "", caption = "Estimator and Test Performance in 5000 simulations of the different estimators and confidence intervals for a binary outcome and completely randomized design.")
```


# Other topics in estimation

## Covariance adjustment: Estimands

In general, simply "controlling for" produces a biased estimator of the ATE
**or** ITT estimand. See for example @lin_agnostic_2013 and @freedman2008rae.
@lin_agnostic_2013 shows how to reduce this bias and, importantly, that this
bias tends to be small as the sample size increases.

## Effects that differ by groups I

If our theory suggests that effects should differ by group, how can we assess evidence for or against such claims?

 - We can **design** for an assessment of this theory by creating a
   block-randomized study --- with blocked defined by the theoretically
   relevant groups.

 - We can **plan** for such an assessment by (1) **pre-registering specific
   subgroup analyses** (whether or not we block on that group in the design
   phase) and (2) making sure to measure group membership during baseline data
   collection pre-treatment


## Effects that differ by groups II
 - If we have not planned ahead, subgroup-specific analyses can be useful as
   explorations but should not be understood as confirmatory: they can too
   easily create problems of testing too many hypotheses thus inflated false
   positive rates.

 - We **should not use groups formed by treatment**. (This is either "mediation
 analysis"  or "conditioning on post-treatment variables" and deserves its own
 module).


## Final thoughts on basics of estimation

- Counterfactual causal estimands are unobserved functions of potential
outcomes.

- Estimators are recipes or functions that use observed data to
learn about an estimand.

- Good estimators produce estimates that are close to the true estimand

- (Connecting estimation with testing) Standard errors of estimators allow us
to calculate confidence intervals and $p$-values. Certain estimators have
larger or smaller (or more or less correct) standard errors.

- You can assess the utility of a chosen estimator for a chosen estimand by simulation.

# Summary of the day

## Summary

 - Statistics helps us **infer** about partially observable counterfactual causal effects
 - We can **hypothesize about those effects** and summarize how much evidence our design and data provide.
 - We can **calculate best guesses** about aggregates of causal effects.
 - We can test hypotheses about the aggregates, too.
 - The statistical properties of our tests and estimators arise from the experimental design, not from a relationship between sample and population.

## References {.allowframebreaks}
